Supplemental Information: Gas exchange in large rivers controlled by hydraulic geometry: implications for remotely sensing gas exchange via the SWOT satellite

### 2021-09-10 supp_v5.Rmd
first ms push with hydraulic geometry stuff

---
title: "Supplemental Information: Gas exchange in large rivers controlled by largest turbulent eddies: implications for remotely sensing gas exchange via SWOT"
author: "CB Brinkerhoff, CJ Gleason, PA Raymond, MH Harlan"
date: "`r Sys.Date()`"
csl: agu_citation_style.csl
output:
  word_document: 
      reference_docx: word-styles-reference.docx
bibliography: References.bib

@@

##Contents
This supplementary information contains 4 texts, 5 figures and 2 tables. Please consult https://github.com/craigbrinkerhoff/RSK600 for all code to build and generate results, figures, and the manuscript.

## Text S1: Deriving two expressions for turbulent kinetic energy production
Below we detail how we arrived at our two expressions for a cross-section-averaged turbulent kinetic energy production (*G*), or equation 3 in the main text. Please see Appendix A for variable descriptions and notation.

@nakagawaTurbulenceOpenChannel1993 outline the standard energy budget for turbulent kinetic energy (TKE), at some distance from the bed *h*, for a 2-dimensional open-channel flow (equation S1). *G* is the TKE produced, $\epsilon$ is the TKE dissipated, $T_d$ is TKE transported from the riverbed via turbulent diffusion, and $V_d$ is TKE transported via viscous diffusion. $T_d$ should always be positive at the free surface, but this is increasingly less true as one moves closer to the riverbed. Since we are interested in the TKE budget at the river surface, $V_d$ can be assumed neglible [@nakagawaTurbulenceOpenChannel1993]. 
$$\mathbf{(S1)} G+T_d+V_d=\epsilon$$

@nakagawaTurbulenceOpenChannel1993 further define *G* and the direct energy dissipation *E*, as averaged over the entire river cross-section and integrated over the entire water column, as equation S2 where *H* is the full water-column depth, *U* is the cross-sectional average flow velocity, and $U_*$ is the friction velocity. $U_*$ is defined by the water surface slope *S*, gravitational acceleration *g* and the channel hydraulic radius $R_h$ as $U_*=\sqrt{gSR_h}$.
$$\mathbf{(S2)} \int_{0}^{H}{Gdh} + \int_{0}^{H}{Edh}=UU_*^2$$

$\int_{0}^{H}{Edh}$ can also be assumed neglible at the free surface [@nakagawaTurbulenceOpenChannel1993]. We then integrate over the full water-column depth *H* to arrive at an expression for *G* at the free surface that reduces to equation S5. Equations S3 to S4 show the algebra.
$$\mathbf{(S3)} \int_{0}^{H}{Gdh} = U\sqrt{gSR_h}^2$$
$$\mathbf{(S4)} GH = UgSR_h$$
$$\mathbf{(S5)} G=gSU\frac{R_h}{H}$$
Alternatively, we can derive another expression for *G* from equation S2. This is provided as equation S8 with the necessary algebra in equations S6-S7.
$$\mathbf{(S6)} \int_{0}^{H}{Gdh}=UU_*^2$$
$$\mathbf{(S7)} GH=UU_*^2$$
$$\mathbf{(S8)} G=\frac{UU_*^2}{H}$$

@@

![Figure S3 Cummulative density functions (CDFs) of river width measurements, as binned by their dataset and and whether or not they meet the 'large river' condition (Rh=H). The dashed black line indicates the minimum river width that the SWOT satellite will observe (100m). This brief analysis indicates that the vast majority of SWOT-observable rivers should meet the 'large river' condition.](cache/k600_theory/swotPlot.jpg)

![Figure S4 Mean river properties (observed k600, width, and slope) for the 47 rivers versus their BIKER performance as defined by NRMSE (see Table S1 for this metric's definition). The dashed black line is the median NRMSE score. The poorly performing rivers are universally the narrowest and steepest of the set.](cache/validation/riverProperties_nrmse.jpg)

![Figure S5 All timeseries of k per river](cache/validation/timeseries_noerr.jpg)

>>>>

---
title: "Supplemental Information: Gas exchange in large rivers controlled by hydraulic geometry: implications for remotely sensing gas exchange via the SWOT satellite"
author: "CB Brinkerhoff, CJ Gleason, CJ Zappa, PA Raymond, MH Harlan"
date: "`r Sys.Date()`"
csl: agu_citation_style.csl
output:
  word_document: 
      reference_docx: word-styles-reference.docx
bibliography: References.bib

@@

##Contents
This supplementary information contains 4 texts, 4 figures and 2 tables. Please consult https://github.com/craigbrinkerhoff/RSK600 for all code to build and generate results, figures, and the manuscript.

## Test S1: Filtering procedure for hydraulics data
To quantify how frequently SWOT-observable rivers are also inefficient river channels (section 2 of main text), we used the dataset of field-measured river hydraulics in @brinkerhoffReconcilingAtaStationAtManyStations2019. That dataset has over 500,000 discrete measurements of river width, velocity, area, and discharge that were made by the United States Geological Survey (USGS) to calibrate streamgauge rating curves. Here, we descirbe how this dataset was filtered down to 171,553 measurements to meet the assumptions in our conceptual model.

First, we removed all measurements missing data, measurements tagged by the USGS as 'poor', measurements with impossible values, or measurements of 0. While this would indicate a dry channel, our hydraulic geometry model necessitates within-bank flow. Likewise, because hydraulic geometry only applies to within-bank flows and not flood events, we had to remove all overbank flows. This was done by first filtering for sites with at least 20 measurements (to build robust estimates of bankfull hydraulics) and then calculating bankfull width and depth as the width or depth with a return period of two years. While the only true way to calculate bankfull hydraulics is manually in the field, this was ovbiously impractical. A two year return period is a standard approximation for determining out-of-bank flow in single-channel meandering rivers and was the method used by @brinkerhoffReconcilingAtaStationAtManyStations2019. We then removed all measurements with a width or depth beyond their respective at-a-station 2 year values.

@@

![Figure S3 All timeseries of k per river](cache/validation/timeseries_noerr.jpg)
